\Large \textbf{Phenotype Enrichment Discovers Phenologs for Disease Modeling in \cel{}}

### 2017-01-05 pea_paper_docs/tex/pea.tex
started writing and improved code base
TEA should run many times faster as a result.

Some numbers may have changed? I can’t figure out if that’s because of
the dictionary I’m using or something else, but I’m worried that there
was a bug prior. This new code is considerably cleaner and it looks
fine to me….

% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8x]{inputenc}
% clean citations
\usepackage{cite}
% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref}
% line numbers
\usepackage[right]{lineno}
% improves typesetting in LaTeX
% \usepackage{microtype}
% degree symbol
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{booktabs}
% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

@@

\title{
  \Large
  \textbf{Phenotype Enrichment Discovers Phenologues for Disease Modeling in \cel{}}
}

@@

\section*{Introduction}

\section*{Results}

\subsection*{Developing PEA and GEA}

\subsection*{Applying the WormBase Enrichment Suite to Discover Disease Phenologues}
\subsubsection*{Obesity}
\subsubsection*{Schizophrenia}
\subsubsection*{Type-2 Diabetes}
\subsubsection*{Crohn's Disease}
\subsubsection*{Prostate Cancer}

\subsection*{Ontology Enrichment as an aid for Screen Design}
\subsubsection*{Enrichment in the Linker Cell Transcriptome}
\subsubsection*{Enrichment in a Linker Cell Screen}

>>>>

% use Unicode characters - try changing the option
% if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8x]{inputenc}
% clean citations
\usepackage{cite}
% hyperref makes references clicky. use \url{www.example.com}
% or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref}
% line numbers
\usepackage[right]{lineno}
% improves typesetting in LaTeX
% \usepackage{microtype}
% degree symbol
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{booktabs}
% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,
            labelsep=period,singlelinecheck=off]{caption}

@@

\newcommand{\hobesity}{957}
\newcommand{\wobesity}{614}

\newcommand{\hschizo}{899}
\newcommand{\wschizo}{433}

\newcommand{\hcrohn}{390}
\newcommand{\wcrohn}{213}

\newcommand{\hprostate}{273}
\newcommand{\wprostate}{273}

\newcommand{\hdiabetes}{396}
\newcommand{\wdiabetes}{224}

\title{
  \Large
  \textbf{Phenotype Enrichment Discovers Phenologs for Disease Modeling in \cel{}}
}

@@

\section*{Introduction}
The last decade has seen an explosion of techniques capable of genome-wide
measurements. Some examples of genome-wide tools include RNA-seq~\cite{} to
measure gene expression, CHIP-seq~\cite{} to measure binding, or genome-wide
methylation profiling to understand epigenetic changes. These tools are capable
of generating large quantities of data. Understanding this data, and generating
hypotheses from it remains challenging. A common approach used to understand
these datasets is to reduce the dimensionality of the data via enrichment
analyses of ontologies, which helps researchers understand what terms are
overrepresented beyond random levels. By analyzing overrepresented terms in
aggregate, researchers can better understand what biological processes were most
affected in a given experiment, and form hypotheses about what is
happening~\cite{}. This approach is limited by what ontologies can be tested for
enrichment. In \cel{}, only the gene ontology (GO) and tissue ontology are
available for testing~\cite{} even though curated gene, tissue and phenotype
ontologies exist. Another limitation is that tissue enrichment testing
is not offered on the same websites as GO enrichment testing, which requires
users to test their data on different websites that may or may not use different
methodologies to detect enrichment.

Another way to use enrichment tools is for evolutionary comparison purposes.
In molecular biology it is often useful to know when a gene is homologous
between two species---that is to say, common by descent---because knowledge of
homology often brings with it knowledge of function~\cite{}. Indeed, many
important gene regulatory networks (GRNs) are conserved between organisms as
highly diverged as nematodes and humans (for example, see~\cite{}). While genes
and GRNs may be conserved between species, their outputs often differ however.
For example, although homeobox genes are known to be important for limb
formation in many animals~\cite{}, these genes do not form limbs in nematodes.
The concept of a phenolog has been put forward to explain relationships between
phenotypes that have the same underlying genetic regulatory network~\cite{}.
Formally, two phenotypes are phenologs of each other if the homologs of the
genes that cause a phenotype in an organism cause a second phenotype in another.

In order to study a clinically relevant disease, an appropriate model has to be
established. A straightforward method towards establishing a disease model in
\cel{} is to link a disease to a causal gene, then to identify the homologous
gene in \cel{} and then to study the function of the genetic homolog to
extrapolate back to humans. However, this method relies on the existence of
known disease genes and requires that the homolog have a phenotype that can be
reliably identified and studied. A fundamentally different way to establish a
disease model in \cel{} would be to identify the phenologs of the disease to be
studied in \cel{}, and to use that phenolog as the basis for screens to identify
genes that are associated with that phenolog. This approach has been
successfully used in the past to make non-obvious links between phenotypes in
different species~\cite{}.

In order to facilitate understanding of large datasets, and to make discovery of
phenologs easier, we have developed a complete enrichment tool suite in WormBase
that allows users to rapidly perform all analyses on curated \cel{} ontologies
using the same methodology for each one. They are located at
% TODO: include links.
We applied our tools towards the unbiased discovery of phenologues of
multigenic, complex diseases including schizophrenia, type-2 diabetes,
crohn's disease and prostate cancer by using genes associated with these
diseases via genome-wide association studies. We also illustrate the utility of
the complete enrichment suite for finding new relationships in complex data by
analyzing a ciliary neuron transcriptome~\cite{}.

\section*{Methods}

\section*{Results}
\subsection*{Developing the WormBase Enrichment Suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{}. We generated a dictionary that included terms with at
least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
(the total number of terms in the dictionary is XXX);\@ and we generated a
dictionary that included terms with at least 100 annotating genes or more and
had a similarity threshold of 0.95 for GEA (the total number of terms in this
dictionary is XXX).\@ Next, we benchmarked the dictionaries on the same gene
sets as TEA and obtained enrichment of all the expected categories. For example,
on a gene set enriched for embryonic muscle genes~\cite{}, the top two enriched
phenotype terms by q-value were `muscle system morphology variant' and `body
wall muscle thick filament variant'; the top two enriched GO terms were
`developmental process' and `contractile fiber'. For all the benchmarking
results, see supplementary information. Having generated and validated our
dictionaries, we proceeded to identify phenologs for several common human diseases.

\subsection*{Applying the WormBase Enrichment Suite}
In order to discover phenologs, we first needed to identify genes that
contribute to a disease in an unbiased manner. One way to discover gene
associations in an unbiased manner is to perform genome-wide association
studies (GWAS) in human populations. Therefore, we used the GWAS NHGRI-EBI
Catalog~\cite{} to identify genes associated with human diseases. We found the
best nematode candidate homologs for these genes using DIOPT~\cite{} and applied
our enrichment suite to each of these gene regulatory networks.

\subsubsection*{Obesity-related traits}
Human genes in Obesity-related traits: 957\\
Worm genes in Obesity-related traits: 614

Obesity-related traits is a category within the GWAS NHGRI-EBI catalog that
pools studies that have measured obesity and other traits associated with
obesity, such as heart rate, physical activity, hormone levels, body composition
and cholesterol levels. Since this category includes many parameters,
we expected there would be many phenologs. GWAS studies have identified
\hobesity{} genes associated with these traits. Using DIOPT, we found
\wobesity{} homologs for these genes, of which XXX had a phenotype.

Top results for obesity-related traits included `acetylcholinesterase inhibitor
response variant', `behavior variant', `neurite morphology variant' and `thin'.
Terms involving locomotion were significantly enriched, as were terms involving
body shape and food consumption ($q<0.1$). Concomitant with these phenologs was
an enrichment in tissues including the \cel{} `tail', `sex organ', and an
eclectic collection of neuronal tissues and cells.

\subsubsection*{Schizophrenia}
Human genes in schizophrenia: 899\\
Worm genes in schizophrenia: 433

Schizophrenia results included `acetylcholinesterase inhibitor response variant',
multiple terms involving vulval formation and induction, cell division defects
(`sister chromatid segregation defective early embryo', `cytokinesis variant')
and terms involving lethality or arrest at multiple stages. Tissue enrichment
showed that the anal depressor and sphincter muscles were significantly
overrepresented, as were neuronal tissues. Somatic gonad, including the
spermatheca and the distal tip cell were also significantly overrepresented in
this dataset.

\subsubsection*{Type-2 Diabetes}
Human genes in diabetes: 396\\
Worm genes in diabetes: 224

The phenolog results for type-2 diabetes included `somatic gonad morphology
variant', and `Q neuroblast lineage migration variant'.

\subsubsection*{Crohn's Disease}
Human genes in Crohn's Disease: 390\\
Worm genes in Crohn's Disease: 213

Crohn's disease in the worm presents traits including `brood size variants',
`P granule defective', `cell fate transformation', `male mating defective' as
well as various defects involving the \cel{} gonad. Reflecting this, TEA showed
that Crohn's associated genes in the worm were enriched in the reproductive
tract and somatic gonad, as well as the dorsal, ventral and lateral nerve cords
and the nerve ring.

\subsubsection*{Prostate Cancer}
Human genes in prostate cancer: 273\\
Worm genes in prostate cancer: 273

Genes associated with prostate cancer in nematodes were associated with cell fate
transformations, adult lethal phenotypes and P granule localization defects,
and enriched tissues included the nerve cords, the sex organs and the tail of the
animal.

\subsection*{Ontology Enrichment as an aid for Screen Design}
An additional use for a tool like PEA would be as a tool to help guide and
design screens to identify genes from an RNA-seq or other genome-wide experiment
for further study. This would be particularly useful in cases when researchers
may not know what phenotype to expect, in which case PEA can guide selection of
a phenotype. Another use case is a scenario where the phenotype under study is
not easy to screen for. By finding phenologs to the phenotype of interest, the
researcher can design an easier screen for genes that affect the phenolog in
question, then re-test genes for the original phenotype of interest.

\subsubsection*{Enrichment in the Ciliary Neuronal Transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes,
we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
they are also present in the male cephalic sensillum and hermaphrodites also have
ciliated neurons. The results are illuminating. PEA reveals that the ciliated
neuron transcriptome is enriched for genes that are typically associated with
chromosome segregation, aneuploidy and spindle defects. This makes sense---cilia
employ microtubules for structural integrity, and microtubules are required for
chromosome segregation. Whereas PEA suggests that this dataset is enriched in
microtubule defects, TEA points at the \cel{} gonad primordium, somatic gonad
and germline precursors as the sites where genes associated with ciliary neurons
are enriched.  \cel{} has a stereotyped cell lineage~\cite{}, and no cellular
division happens after a certain point in post-embryonic development, with the
exception of the germline.

\section*{Conclusions}
